---
title: "9: Data Visualization"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## LESSON OBJECTIVES
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrient_Phys_PeterPaul_Processed.csv")
PeterPaul.nutrients.gathered <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
PeterPaul.chem.nutrients.summaries <- read.csv("./Data/Processed/NTL-LTER_Lake_Summaries_PeterPaul_Processed.csv")

EPAair <- read.csv("./Data/Processed/RM_EPAair_O3_PM25_NC1718_Processed.csv") 

#Taylor's fix for date
EPAair <- EPAair %>%
  unite(Date, Y, m, d)
EPAair$Date <- ymd(EPAair$Date)

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")

dim(EPAair)
class(EPAair$Date)


PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")
View(PeterPaul.chem.nutrients)
```

## GGPLOT

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 

### Geoms

Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

### Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

### Plotting continuous variables over time: Scatterplot
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
ggplot(EPAair, aes(x = Date, y = Ozone)) + 
  geom_point()

O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot)

# Fix this code
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone, color = Ozone)) #make color the Ozone
print(O3plot2)

# Add additional variables - I missed this
PMplot <- 
  ggplot(EPAair, aes(x = month, y = PM2.5, shape = as.factor(year), color = Site.Name)) +
  geom_point()
print(PMplot)

# Separate plot with facets  - I missed this
PMplot.faceted <-
  ggplot(EPAair, aes(x = month, y = PM2.5, shape = as.factor(year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3) 
print(PMplot.faceted) #will have common legend

# - I missed this
PMplot.faceted2 <-
  ggplot(EPAair, aes(x = month, y = PM2.5)) +
  geom_point() +
  facet_grid(Site.Name ~ year)
print(PMplot.faceted2)

# Filter dataset within plot building
O3plot.Blackstone <- 
  ggplot(subset(EPAair, Site.Name == "Blackstone"), aes(x = Date, y = Ozone)) + #subset can take name
  geom_point() +
  geom_line() #Added layer, line will only go from one entry to another - day 1 to 2 to 3 (but if it was an entry every month, that wouldn't be useful to)
print(O3plot.Blackstone)

# Exercise: build your own scatterplots of PeterPaul.chem.nutrients

# 1. 
# Plot surface temperatures by day of  year. 
# Color your points by year, and facet by lake in two rows.

tn_plot.PeterPaul <-
  ggplot(subset(PeterPaul.chem.nutrients, depth = 0),
         aes(x = daynum, y = temperature_C, color = year4))+
  geom_point()+
  facet_wrap(vars(lakename), nrow = 2) #put them into 2 rows vs. default of 2 columns


#2. 
# Plot temperature by date. Color your points by depth.
# Change the size of your point to 0.5
  ggplot(PeterPaul.chem.nutrients,
         aes(x = sampledate, y = temperature_C, color = depth))+
  geom_point(size = 0.5)

```
### Plotting the relationship between two continuous variables: Scatterplot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO) #this makes no sense! there's a couple irrational (not true) outliers
#zoom into area of interest

# Adjust axes
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) + #adding limits so we have a better view
  ylim(0, 20)
print(lightvsDOfixed) #shows lots of light at surface, less at depth; less DO at low irradiance levels, and more instances of higher DO at surface

# Depth in the fields of limnology and oceanography is on a reverse scale, so we can see data differently
tempvsdepth <- 
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse()
print(tempvsdepth) #more fluctuation at surface, colder at depth
 
##where we stopped 2/7/19
NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) +
  geom_point() +
  geom_smooth(method = lm) + #telling it to be a linear model method (otherwise it just uses smooth function for fit line); the line also includes a 95% confidence interval
  geom_abline(aes(slope = 16, intercept = 0), lty = 2) #automatically will make it black straight line, line type has to go outside aes
print(NvsP)
#this shows ratio of 16:1; the line shows most points above; this is expected for great lakes

# Exercise: Plot relationships between air quality measurements

# 1. 
# Plot AQI values for ozone by PM2.5, colored by site. 
View(EPAair)
EPA.O3.PM2.5.sites <- 
  ggplot(EPAair, aes(x = PM2.5, y = Ozone)) +
  geom_point(color = Site.Name) + #you can put aes in geom_point, but may need to repeat
  geom_smooth(method = lm) #both include fit lines for each variable 
print(EPA.O3.PM2.5.sites) #triple oak doesn't show up, but there's no ozone data, so it won't show up in graph
#scale color discrete to specify colors, it's its own layer

# Add a line of best fit for the linear regression of these variables.

  
```

### Plotting continuous vs. categorical variables

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
 # Barplot + error bars
PeterPaul.nutrient.summaries <- PeterPaul.nutrients.gathered %>%
  group_by(lakename, nutrient) %>% #if you want to group_by more than one column, they have to be in same fun
  summarise(sd = sd(concentration), 
            mean = mean(concentration)) 

Nutrientplot <- 
  ggplot(PeterPaul.nutrients.gathered) +
  geom_bar(aes(x = lakename, y = concentration, fill = as.factor(nutrient)), # why did we use fill? color will only color outside of line
          position = "dodge", stat = "summary", fun.y = "mean")             # what's happening here? #position = dodge computes mean for each nut, instead of making single bar for each lake; stat = summary & fun.y = mean telling it to do summary of the mean (though it does choose mean by default)
print(Nutrientplot)

class(PeterPaul.nutrients.gathered)

Nutrientplot2 <- 
  ggplot(PeterPaul.nutrient.summaries, aes(x = lakename, y = mean, fill = as.factor(nutrient))) + #splitting it up by nut
  geom_bar(stat = "identity", position = "dodge") + # what does the stat command do? bar chart requires a command specifying stat; "dodged" out different bars
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd),
                position = "dodge") # how do we specify error bars?
                 #we are using these means that we've already computed
print(Nutrientplot2) #but this is a crazy way to show nut concns: no error bars makes more sense; we're moving away from bar plots w/ error bars

# Are there more effective ways to produce summary stats for categories?

# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"? fill only fills in the boxes, but makes all points black 
print(Nutrientplot3)

# Dot plot = wide spots are where there's more measurements / entries
Nutrientplot4 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient), binaxis = "y", binwidth = 1, #binwidth 
               stackdir = "center", position = "dodge") #stack direction
print(Nutrientplot4) #if the data weren't as widely distrib, we'd get better more informative graph

# Violin plot looks more like a violin when points are closer together
Nutrientplot5 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #
print(Nutrientplot5)

# Frequency polygons
# Using a tidy dataset
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "black") + #if you don't specify colors, it'll just make black lines
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "darkgray") +
  geom_freqpoly(aes(x = no23), color = "gray") +
  geom_freqpoly(aes(x = po4), color = "blue") #one layer for each nut
print(Nutrientplot6)

# Using a gathered dataset
Nutrientplot7 <-   
  ggplot(PeterPaul.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient), binwidth = 1000)
print(Nutrientplot7)

# Exercise: Plot distributions of AQI values for EPAair

# 1. 
# Create a bar chart plus standard deviation error bars for PM2.5, divided by year. 
class(EPAair$Date) #need to mutate this file to add month and year

EPA.PM2.5.years <- EPAair %>%
  group_by(Site.Name, year) %>%
  summarise(sd = (sd(PM2.5, na.rm = TRUE)), 
            mean = (mean(PM2.5, na.rm = TRUE)))

  ggplot(EPA.PM2.5.years, aes( x = as.factor(year), y= mean, fill = Site.Name)) +
  geom_bar(position = "dodge", stat = "identity") +
    geom_errorbar(aes(ymin = mean - sd, ymax = mean +sd), position = "dodge") 
  #geom_errorbar wants numbers, be careful
  
# Create separate bars for each site. 

# 2. 
# Create a new plot that better depicts the distribution of PM2.5 concentrations. 
# Divide your graph by year and site.

```
