---
title: "11: Generalized Linear Models"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## LESSON OBJECTIVES
1. Describe the components of the generalized linear model (GLM)
2. Apply special cases of the GLM to real datasets
3. Interpret and report the results of GLMs in publication-style formats

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
class(PeterPaul.chem.nutrients$sampledate)
# Set date to date format
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, 
                                               format = "%Y-%m-%d")

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## SIMPLE AND MULTIPLE LINEAR REGRESSION
The linear regression, like the t-test and ANOVA, is a special case of the **generalized linear model** (GLM). A linear regression is comprised of a continuous response variable, plus a combination of 1+ continuous response variables (plus the error term). The deterministic portion of the equation describes the response variable as lying on a straight line, with an intercept and a slope term. The equation is thus a typical algebraic expression: 
$$ y = \alpha + \beta*x + \epsilon $$

The goal for the linear regression is to find a **line of best fit**, which is the line drawn through the bivariate space that minimizes the total distance of points from the line. This is also called a "least squares" regression. The remainder of the variance not explained by the model is called the **residual error.** 

The linear regression will test the null hypotheses that

1. The intercept (alpha) is equal to zero.
2. The slope (beta) is equal to zero

Whether or not we care about the result of each of these tested hypotheses will depend on our research question. Sometimes, the test for the intercept will be of interest, and sometimes it will not.

Important components of the linear regression are the correlation and the R-squared value. The **correlation** is a number between -1 and 1, describing the relationship between the variables. Correlations close to -1 represent strong negative correlations, correlations close to zero represent weak correlations, and correlations close to 1 represent strong positive correlations. The **R-squared value** is the correlation squared, becoming a number between 0 and 1. The R-squared value describes the percent of variance accounted for by the explanatory variables. 

### Simple Linear Regression
For the NTL-LTER dataset, can we predict irradiance (light level) from depth?
```{r}
irradiance.regression <- lm(PeterPaul.chem.nutrients$irradianceWater ~ PeterPaul.chem.nutrients$depth)
# another way to format the lm function
irradiance.regression <- lm(data = PeterPaul.chem.nutrients, irradianceWater ~ depth)
summary(irradiance.regression) #y var is on left, explan var is on right
#can specify df and then give it both var
# the alpha (irrad intercept) is about 487 (signif dif than 0); the Rsq value is only explaining 31% of 
#look for coeff and beta term 
#adj Rsq will penalize for having more variables

# Correlation
cor.test(PeterPaul.chem.nutrients$irradianceWater, PeterPaul.chem.nutrients$depth)
```
Question: How would you report the results of this test (overall findings and report of statistical output)?

> ANSWER: R2 = 0.31, pval < 0.0001 (use the one form F-stat right here), df = 15449, also this is a linear regression test (where we wanna know R2 and pval).
The R2 value is only about 31%, so
we have high number of df so we have more wiggle room (since we have so many observations)
CI:
Irrad was neg corr w/ depth. 
put linear regr eqn in figure (use geom_text)

So, we see there is a significant negative correlation between irradiance and depth (lower light levels at greater depths), and that this model explains about 31 % of the total variance in irradiance. Let's visualize this relationship and the model itself. 

An exploratory option to visualize the model fit is to use the function `plot`. This function will return four graphs, which are intended only for checking the fit of the model and not for communicating results. The plots that are returned are: 

1. **Residuals vs. Fitted.** The value predicted by the line of best fit is the fitted value, and the residual is the distance of that actual value from the predicted value. By definition, there will be a balance of positive and negative residuals. Watch for drastic asymmetry from side to side or a marked departure from zero for the red line - these are signs of a poor model fit.

2. **Normal Q-Q.** The points should fall close to the 1:1 line. We often see departures from 1:1 at the high and low ends of the dataset, which could be outliers. 

3. **Scale-Location.** Similar to the residuals vs. fitted graph, this will graph the squared standardized residuals by the fitted values. 

4. **Residuals vs. Leverage.** This graph will display potential outliers. The values that fall outside the dashed red lines (Cook's distance) are outliers for the model. Watch for drastic departures of the solid red line from horizontal - this is a sign of a poor model fit.

```{r, fig.height = 3, fig.width = 4}
plot(irradiance.regression) #to learn more about data, looking at resid vs fitted;resid vs leverage, etc
```

The option best suited for communicating findings is to plot the explanatory and response variables as a scatterplot. 

```{r, fig.height = 3, fig.width = 4}
# Plot the regression
irradiancebydepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) + 
  ylim(0, 2000) + #to zoom into plot, bc there was a crazy outlier
  geom_point() 
print(irradiancebydepth) 
#this wasn't the best way to know our data (a linear reg doesn't make sense, it's exponential); so it's better to use log transform
```

Given the distribution of irradiance values, we don't have a linear relationship between x and y in this case. Let's try log-transforming the irradiance values.

```{r, fig.height = 3, fig.width = 4}
PeterPaul.chem.nutrients <- filter(PeterPaul.chem.nutrients, irradianceWater != 0) #removing zeros bc ya can't log them
irradiance.regression2 <- lm(data = PeterPaul.chem.nutrients, log(irradianceWater) ~ depth)
summary(irradiance.regression2)
plot(irradiance.regression2)
#log(irradiance) = 6.2 - 0.74(depth) + E is the eqn for

# Add a line and standard error for the linear regression
irradiancebydepth2 <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
  geom_smooth(method = "lm") + #the line may show that there may not be cleanest relationship at certian depths
  scale_y_log10() +
  geom_point() 
print(irradiancebydepth2) 
#same aes, but add scale y log10

# SE can also be removed
irradiancebydepth2 <- 
    ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
    geom_point() +
    scale_y_log10() +
    geom_smooth(method = 'lm', se = FALSE, color = "pink") #can remove std error
print(irradiancebydepth2)

# Make the graph attractive

irradiancebydepth2 <- 
    ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater, color = lakename, fill = )) +
    geom_point(alpha = 0.8) + # makes it 20% transparent
    scale_y_log10() +
  labs(x = "Depth, m", y = "Water Irradiance, (units)")
    geom_smooth(method = 'lm', se = FALSE, color = "pink") + #can remove std error
#legend(legend.position = "right")
      scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12)) +
      #geom_text(put your eqn here somehow) look up annotation layer
print(irradiancebydepth2)
```

### Non-parametric equivalent: Spearman's Rho
As with the t-test and ANOVA, there is a nonparametric variant to the linear regression. The **Spearman's rho** test has the advantage of not depending on the normal distribution, but this test is not as robust as the linear regression.

``` {r}
cor.test(PeterPaul.chem.nutrients$irradianceWater, PeterPaul.chem.nutrients$depth, 
         method = "spearman", exact = FALSE)
```

### Multiple Regression
It is possible, and often useful, to consider multiple continuous explanatory variables at a time in a linear regression. For example, total phosphorus concentration could be dependent on depth and dissolved oxygen concentration: 

``` {r, fig.height = 3, fig.width = 4}
TPregression <- lm(data = PeterPaul.chem.nutrients, tp_ug ~ depth + dissolvedOxygen) #response on left, explan on right
summary(TPregression)  #we see that TP = 6 +1.5(depth) + 0.94 (DO) + E; we're only explaining 8% of var

TPplot <- ggplot(PeterPaul.chem.nutrients, 
                 aes(x = dissolvedOxygen, y = tp_ug, color = depth)) +
  geom_point() +
  xlim(0, 20)
print(TPplot)

TPplot <- ggplot(PeterPaul.chem.nutrients, 
                 aes(x = depth, y = dissolvedOxygen, color = depth)) +
  geom_point() +
  xlim(0, 20)
print(TPplot)
```
### Correlation Plots
We can also make exploratory plots of several continuous data points to determine possible relationships, as well as covariance among explanatory variables. 

```{r, fig.height = 3, fig.width = 4}
#install.packages("corrplot") #to think about relationships btwn contin data
library(corrplot)
PeterPaulnutrients <- 
  PeterPaul.chem.nutrients %>%
  select(tn_ug:po4) %>% #so you are only working w/ certain columns, bc na.omit will take out na in entire dataset otherwise
  na.omit()
PeterPaulCorr <- cor(PeterPaulnutrients)
corrplot(PeterPaulCorr, method = "ellipse") #shows interesting stuff about P and PO4 (is there organic P or what?)
corrplot.mixed(PeterPaulCorr, upper = "ellipse") #cool informative way to show matrix...but you must you know your data to make sure it is useful/makes sense
#(be wary of throwing so many corr together that you get truly weird stuff - spaghetti at the wall to see what sticks actually just leaves a mess)
```
### AIC to select variables

However, it is possible to over-parameterize a linear model. Adding additional explanatory variables takes away degrees of freedom, and if explanatory variables co-vary the interpretation can become overly complicated. Remember, an ideal statistical model balances simplicity and explanatory power! To help with this tradeoff, we can use the **Akaike's Information Criterion (AIC)** to compute a stepwise regression that either adds explanatory variables from the bottom up or removes explanatory variables from a full set of suggested options. The smaller the AIC value, the better. 

Let's say we want to know which explanatory variables will allow us to best predict total phosphorus concentrations. Potential explanatory variables from the dataset could include depth, dissolved oxygen, temperature, PAR, total N concentration, and phosphate concentration. 

```{r}
PeterPaul.naomit <- na.omit(PeterPaul.chem.nutrients) #only have 242 obs now
TPAIC <- lm(data = PeterPaul.naomit, tp_ug ~ depth + dissolvedOxygen + 
              temperature_C + tn_ug + po4)
step(TPAIC) # will run it w/out certain var stepwise, to get better best fit model(can tell it to build from ground up increasing.step or something) ; the lower AIC val we have the better the best fit is; so we look for lowest AIC
TPmodel <- lm(data = PeterPaul.naomit, tp_ug ~ temperature_C + tn_ug)
summary(TPmodel) #now we've explained 54% (rather than only 8% like before). But be careful bc this could lead to laziness which could lead to bad assumptions

```

## ANCOVA
Analysis of Covariance consists of a prediction of a continuous response variable by both continuous and categorical explanatory variables. We set this up in R with the `lm` function, just like prior applications in this lesson. 

Let's say we wanted to predict total nitrogen concentrations by depth and by lake, similarly to what we did with a two-way ANOVA for depth ID and lake. 

```{r, fig.height = 3, fig.width = 4}
# main effects
TNancova.main <- lm(data = PeterPaul.chem.nutrients, tn_ug ~ lakename + depth) #main effect (we can tell by the + sign) using depth as continuous rather than categorical since we have one categ already
summary(TNancova.main) #TN = 353 + 0(depth)for Paul + 136for Peter + E, only explaining 9% of variance

# interaction effects
TNancova.interaction <- lm(data = PeterPaul.chem.nutrients, tn_ug ~ lakename * depth)
summary(TNancova.interaction)


TNplot <- ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = tn_ug, color = lakename)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + #if no color specified, it gave us 2 lines, if color = "something" it gives us 1 line
  xlim(0, 10)
print(TNplot)

# Make the graph attractive
```

