---
title: "4: Coding Basics"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## LESSON OBJECTIVES
1. Develop familiarity with the form and function of the RStudio interface.
2. Apply basic functionality of R
3. Evaluate how basic practies in R contribute to best management practices for data analysis

## DATA TYPES IN R
R treats objects differently based on their characteristics. For more information, please see: https://www.statmethods.net/input/datatypes.html. 

* **Vectors** 1 dimensional structure that contains elements of the same type.

* **Matrices** 2 dimensional structure that contains elements of the same type.

* **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

* **Lists** Ordered collection of elements that can have different modes.

* **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (# at beginning of line, ---- at end)
```{r}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector
vector1
```

```{r}
vector2 <- c("one","two","three") # character vector has to have quotes
vector2
```

```{r}
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
vector3
```

```{r}
vector1[3] #gave us third number of vector1 [this is mattrix subsetting]
```

```{r}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector
vector1
vector2 <- c("one","two","three") # character vector
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
vector3

vector1[3] # 
# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # numbers 1 through 20, w/ 5 rows & 4 columns
matrix1
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) #
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE,
 # put a comma, and then return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "cuatro", "cinco"), 
                                  c("un", "deux", "trois", "cat"))) #
matrix3
```

```{r}
matrix1[4, ] #showed the numbers in the fourth row [row, column]
matrix1[ , 3] #showed the numbers in the third column
matrix1[c(12, 14)] #take subset of matrix1, just the 12th and 14th numbers, the c means concatenate
matrix1[c(12:14)] #show elements between 12 and 14
matrix1[2:4, 1:3] #rows 2 thru 4, columns 1 thru 3

cells <- c(1, 26, 24, 68)  
rnames <- c("R1", "R2") #create an object called rnames
cnames <- c("C1", "C2") 
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # the nrow and ncol has to be a multiple of the object; if this on isn't a multiple of 2, it'll just start over w/ 1
matrix4

# Lists ---- 
list1 <- list(name = "Fred", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1 
list1[[2]]
#can't name object without naming the function ; means it'll act like a separate line of code
#and list can be dif modes of data
#there will be dif ways to get to objects in the list,

# Data Frames ----
d <- c(1, 2, 3, 4) # What type of vector? numerical
e <- c("red", "white", "red", NA) # What type of vector?, character, stored as an NA
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector? logical statement
dataframe1 <- data.frame(d,e,f) #combine d, e, and f
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # 

dataframe1[1:2] # showing columns 1 and 2
dataframe1[c("ID","Passed")] #showing dataframe1 and naming columns ID and Passed
dataframe1$Color # the $ shows us different elements that we can chose from, and the options

```
QUESTION: How do the different types of data appear in the Environment tab?

> ANSWER: 
Only objects that have been defined will show up in Environment. They can be displayed in a List, or a Grid.

QUESTION: In the R chunk below, write "dataframe1$". Press `tab` after you type the dollar sign. What happens?

> ANSWER: 
The $ shows different elements that we can choose from.

QUESTION: What happens when a comment in R is followed by "----"?

> ANSWER:
That section becomes collapsible.

Advanced: Sequential section headers can be created by using at least four -, =, and # characters.

## LIBRARIES/PACKAGES

The Packages tab in the notebook stores the packages that you have saved in your system. A checkmark next to each package indicates whether the package has been loaded into your current R session. Given that R is an open source software, users can create packages that have specific functionalities, with complicated code "packaged" into a simple commands.

If you want to use a specific package that is not in your libaray already, you need to install it. You can do this in two ways: 

1. Click the install button in the packages tab. Type the package name, which should autocomplete below (case matters). Make sure to check "intall dependencies," which will also install packages that your new package uses. 

2. Type `install.packages("packagename")` into your R chunk or console. It will then appear in your packages list. You only need to do this once. 

If a library is already installed, you will need to load it every session. You can do this in two ways: 

1. Click the box next to the package name in the Packages tab. 

2. Type `library(packagename)` into your R chunk or console.

**Tips and troubleshooting**

* You may be asked to restart R when installing or updating packages. Feel free to say no, as this will obviously slow your progress. However, if the functionality of your new package isn't working properly, try restarting R as a first step. 

* If asked "Do you want to install from sources the packages which needs compilation?", type `yes` into the console. 
* You should only install packages once on your machine. If you store `install.packages` in your R chunks/scripts, comment these lines out, as below. 

* Update your packages regularly! 


```{r}
# We will use the packages dplyr and ggplot2 regularly. 
#install.packages("dplyr") # comment out install commands, use only when needed and re-comment
#install.packages("ggplot2")

library(dplyr)
library(ggplot2)

# Some packages are umbrellas under which other packages are loaded
#install.packages(tidyverse)
library(tidyverse)
```
What happens in the console when you load a package?

> ANSWER: 
It either gives you an error (if it didn't work), a warning (if there's something that )
## FUNCTIONS

You've had some practice with functions with the simple commands you've entered in this lesson and the one previous. the basic form of a function is `functionname()`, and the packages we will use in this class will use these basic forms. However, there may be situations when you will want to create your own function. Below is a description of how to write functions through the metaphor of creating a recipe. Credit for this goes to Isabella R. Ghement (@IsabellaGhement on Twitter). 

Writing a function is like writing a recipe. Your function will need a recipe name (functionname). Your recipe ingredients will go inside the parentheses. The recipe steps and end product go inside the curly brackets.
```{r} 
functionname <- function(){
  
}
```

A single ingredient recipe: 
```{r}
# Write the recipe
recipe1 <- function(x){
  mix <- x*2 # x is the ingredient
  return(mix) #mix is the instructions
}

# Bake the recipe
simplemeal <- recipe1(5)

# Serve the recipe
simplemeal
```

Two single ingredient recipes, baked at the same time: 
```{r}
recipe2 <- function(x){
  mix1 <- x*2
  mix2 <- x/2
  return(list(mix1 = mix1, #comma indicates we continue onto the next line
              mix2 = mix2))
}

doublesimplemeal <- recipe2(6)
doublesimplemeal
```
Two double ingredient recipes, baked at the same time: 
```{r}
recipe3 <- function(x, f){
  mix1 <- x*f
  mix2 <- x/f
  return(list(mix1 = mix1, #comma indicates we continue onto the next line
              mix2 = mix2))
}

doublecomplexmeal <- recipe3(x = 5, f = 2) #using the definitions of the ingredients is more helpful
doublecomplexmeal #serve the recipe
doublecomplexmeal$mix1 #the $ means extract phrase that follows
```
Make a recipe based on the ingredients you have
```{r}
recipe4 <- function(x) { #named recipe, using that assignment for it
  if(x < 3) { #if X is less than 3, then multiply it times 2
    x*2
  } 
  else { #else = if not, then divide x by 2
    x/2
  }
}

recipe5 <- function(x) {
  if(x < 3) {
    x*2
  } 
  else if (x > 3) {
    x/2
  }
  else { #if not that previous stuff, then just give me x
    x
  }
}
meal <- recipe4(4); meal
meal2 <- recipe4(2); meal2
meal3 <- recipe5(3); meal3

recipe6 <- function(x){
  ifelse(x<3, x*2, x/2)
  
}

meal4 <- recipe6(4); meal4
meal5 <- recipe6(2); meal4
```

```{r}
meal5 <- recipe6(2); meal5
```

