---
output:
  pdf_document: default
  html_document: default
---
# Environmental Data Analytics Coding Challenge #1: Data Exploration

## The following code explores the EPA ECOTOX database entries for neonicotinoid mortality.
## The code contains numerous mistakes and errors, which you are tasked with fixing.
## Instructions listed in comments throughout the script.

#### Setup ----
```{r}
getwd()
# it will automatically got to where the file is; use relative file path: project wd
```

```{r}
library("tidyverse")
```

```{r}
ecotox.neonic <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Mortality_raw.csv")
# to go back into folder structure, use ../ then tab (to go forward into folder, use ./ then tab)
```

#### Basic Data Summaries ----
head(ecotox.neonic, )
summary(ecotox.neonic%Chemical.Name)
summary(ecotox.neonic%Pub..Year)
```{r}
head(ecotox.neonic, 5) #tell it how many lines to show
summary(ecotox.neonic$Chemical.Name) #have to use $ not %
summary(ecotox.neonic$Pub..Year) 
```

# Fix formatting of column names (spaces originally present were turned to periods upon import)
colnames(ecotox.neonic)[8:12] <- c(Duration, Conc.Type, Conc.Mean, Conc.Units, Pub.Year)
```{r}
#lets look at dimensions first
dim(ecotox.neonic) #1283observations  13variables
str(ecotox.neonic) #whoa its a dataframe
colnames(ecotox.neonic) <- c("CAS.n", "ChemicalName", "SpeciesName",
                             "CommonName", "Effect", "Measurement",
                             "Endpoint", "Duration", "Conc.Type", 
                             "Conc.Mean", "Conc.Units", "Pub.Year",
                             "Citation") 
#rename columns, use quotes ;  if I just wanna show SOME of the columns? have to specify which number column: ex: columns 8 thru 12 is [8:12] 
```

# Plot histogram of counts of publication years
ggplot(ecotox.neonic, aes(x = Pub.Year)) 
  geom_histogram()
```{r}
ggplot(ecotox.neonic) + #make sure it knows to continue
  geom_histogram(aes(x = Pub.Year), bins = 50) #the histogram has to have an x-axis
```

# Plot histogram of counts chemical names
# hint: what is the class of Chemical.Name? There are two options for a solution.
ggplot(ecotox.neonic, x = Chemical.Name) +
  geom_histogram()
```{r}
ggplot(ecotox.neonic, aes(x = ChemicalName)) +
  geom_bar() #make sure it's the name that actually is
#histogram wouldn't work with discrete variable, that deosn't make sense at all; use geom_bar
```

# Plot frequency polygon of publication years divided by chemical name
# Define colors as something other than ggplot default 
ggplot(ecotox.neonic) +
  geom_freqpoly(aes(x = Pub.Year, color = Chemical.Name) +
  theme(legend.position = "right")
```{r}
ggplot(ecotox.neonic) +
  geom_freqpoly(aes(x = Pub.Year, color = ChemicalName), binwidth = 10) +
  scale_color_manual(values = c("blue", "red", "yellow", "orange", "purple", "pink", "green", "turquoise", "brown"))
  theme(legend.position = "right")
```

