---
title: 'Assignment 6: Generalized Linear Models'
author: "Laurie Muzzy"
geometry: margin=2.54cm
output:
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio. 
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
You will need to have the correct software installed to do this (see Software Installation Guide)
Press the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Assignments folder.
8. After Knitting, please submit the completed exercise (PDF file) to the dropbox in Sakai. Please add your last name into the file name (e.g., "Salk_A06_GLMs.pdf") prior to submission.

The completed exercise is due on Tuesday, 26 February, 2019 before class begins.

## Set up your session 
1. Set up your session. Upload the EPA Ecotox dataset for Neonicotinoids and the NTL-LTER raw data file for chemistry/physics. 

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
library(tidyverse)
getwd() 
ECOTOX_Neonic <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Mortality_raw.csv", header = TRUE) #header = TRUE to make sure there's no spaces in column names
library(readr)
NTL_LTER_Lake_ChemistryPhysics_Raw <- read_csv("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
#NTL-LTER_Lake_ChemistryPhysics_Raw <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

#2
A6theme <- theme_gray(base_size = 13)
theme(axis.text = element_text(color = "black"), legend.position = "right")
theme_set(A6theme)
```

## Neonicotinoids test
Research question: Were studies on various neonicotinoid chemicals conducted in different years? 

3. Generate a line of code to determine how many different chemicals are listed in the Chemical.Name column.

4. Are the publication years associated with each chemical well-approximated by a normal distribution? Run the appropriate test and also generate a frequency polygon to illustrate the distribution of counts for each year, divided by chemical name. Bonus points if you can generate the results of your test from a pipe function. No need to make this graph pretty.

5. Is there equal variance among the publication years for each chemical? Hint: var.test is not the correct function.

```{r studies in different years}
#3 how many chemicals are listed 
summary(ECOTOX_Neonic$Chemical.Name) #(9 chemicals)
class(ECOTOX_Neonic$Pub..Year) #integer

#4  see if it's a normal distr 
#not numeric, so need other test like ANOVA

Chem.Name <- function(N) {ECOTOX_Neonic %>%
    filter(Chemical.Name == 'Acetamiprid') %>%
    pull(Pub..Year) %>%
    shapiro.test()
}

Chem.Name
Ecotox.PubYr.norm <- ggplot(ECOTOX_Neonic) +
geom_freqpoly(aes(x = Pub..Year, color = Chemical.Name), stat = "count") +
  labs(x = "year of publication", y = "number of publications")
print(Ecotox.PubYr.norm)

#5 equal var in pub yrs for each chemical?

bartlett.test(ECOTOX_Neonic$Pub..Year ~ ECOTOX_Neonic$Chemical.Name) 
#Bartlett's K-squared = 139.59, df = 8, p-value < 2.2e-16

# p <0.0001, so we can reject the null; the variance is not the same for all the chemicals.
```

6. Based on your results, which test would you choose to run to answer your research question?

> ANSWER: "Were studies on various neonicotinoid chemicals conducted in different years?"
Kruskal-Wallis test, because it compares multiple groups and it's nonparametric.

7. Run this test below. 

8. Generate a boxplot representing the range of publication years for each chemical. Adjust your graph to make it pretty.
```{r pub years for chemicals}
#7 test for studies of dif chemicals performed in dif years
#response ~ explanatory
range(ECOTOX_Neonic$Pub..Year)
summary(ECOTOX_Neonic$Chemical.Name)
Chem.PubYr.kruskal <- kruskal.test(Pub..Year ~ Chemical.Name, ECOTOX_Neonic)
Chem.PubYr.kruskal #Kruskal-Wallis chi-squared = 134.15, df = 8, p-value < 2.2e-16

#8 boxplot of range of pub years for each chemical
#not informative enough: need better x axis, can't figure out units or numbers
Ecotox.PubYr.Chemicals <- ggplot(ECOTOX_Neonic, aes(stat = "count", y = Pub..Year )) + # x = count just puts "count" on x-axis, not helpful
geom_boxplot(aes(fill = Chemical.Name), position = "dodge") + 
#labs(x = "number of publications", y = "publication year", title = "Publications on Neonicotinoids, 1982-2018") +
theme(legend.position = "right")
print(Ecotox.PubYr.Chemicals)

```


9. How would you summarize the conclusion of your analysis? Include a sentence summarizing your findings and include the results of your test in parentheses at the end of the sentence. 

> ANSWER: Used Kruskal test, p-val <0.05, indicating significant difference between the amount of publications for the different chemicals. (results: Kruskal-Wallis chi-squared = 134.15, df = 8, p-value < 2.2e-16)

## NTL-LTER test
Research question: What is the best set of predictors for lake temperatures in July across the monitoring period at the North Temperate Lakes LTER? 

11. Wrangle your NTL-LTER dataset with a pipe function so that it contains only the following criteria: 

* Only dates in July (hint: use the daynum column). No need to consider leap years.
* Only the columns: lakename, year4, daynum, depth, temperature_C
* Only complete cases (i.e., remove NAs)

12. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature. Run a multiple regression on the recommended set of variables. 

```{r}
#11 : dates in July: lakename, year4, daynum, depth, temperature_C, remove NAs (na.omit but only after pipe)
Lake.July.temps <- NTL_LTER_Lake_ChemistryPhysics_Raw %>%
filter(daynum >= 182 & daynum <= 212) %>%
select(lakename, year4, daynum, depth, temperature_C) %>%
na.omit()

#12 AIC

#Correlations close to -1 represent strong negative correlations, correlations close to zero represent weak correlations, and correlations close to 1 represent strong positive correlations. The **R-squared value** is the correlation squared, becoming a number between 0 and 1. The R-squared value describes the percent of variance accounted for by the explanatory variables.
Lake.July.temps.AIC <- lm(data = Lake.July.temps, temperature_C ~ depth + daynum + year4)
step(Lake.July.temps.AIC)
Lake.July.temps.model <- lm(data = Lake.July.temps, temperature_C ~ year4 + daynum)
step(Lake.July.temps.model)
summary(Lake.July.temps.model) #Residual standard error: 7.489 on 9719 degrees of freedom Multiple R-squared:  0.002489,	Adjusted R-squared:  0.002284 F-statistic: 12.13 on 2 and 9719 DF,  p-value: 5.503e-06
#weak correlation: only 0.2% of variance is accounted for by explan var

Lake.July.temps.regression <- lm(data = Lake.July.temps, temperature_C ~ year4 + daynum)
summary(Lake.July.temps.regression)

Lake.July.temps.plot1 <- ggplot(Lake.July.temps, 
                 aes(x = temperature_C, y = year4, color = daynum)) +
  geom_point(size = 1) 
print(Lake.July.temps.plot1)

Lake.July.temps.plot2 <- ggplot(Lake.July.temps, 
                 aes(x = daynum, y = temperature_C, color = year4)) +
  geom_point(size = 1)
print(Lake.July.temps.plot2)
```

13. What is the final linear equation to predict temperature from your multiple regression? How much of the observed variance does this model explain?

> ANSWER: temperature_C = 0.36 + 0.002(year4) + 0.04(daynum) + 16.9(E). This model only explains 0.2% of the variance, which is terrible.
 
14. Run an interaction effects ANCOVA to predict temperature based on depth and lakename from the same wrangled dataset.

```{r}
#14 lm

Lake.July.temps.ancova <- lm(data = Lake.July.temps, temperature_C ~ lakename + depth)
summary(Lake.July.temps.ancova)

```

15. Is there an interaction between depth and lakename? How much variance in the temperature observations does this explain?

> ANSWER: There appears to be an interaction between depth and lakename (which makes sense: lakes are probably going to have different depths). It explains about 78% of the variance.

16. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r}
#16 x=depth y=temp ?

Lakes.Temp.by.depth <- ggplot(Lake.July.temps, aes(x = depth, y = temperature_C), color = depth) +
  theme_bw() +
  geom_point(alpha = 0.5, size = 0.2, color = "gray") +
  ylim(0,35) +
  geom_smooth(aes(color = lakename), method = "lm", se = FALSE, size = 0.5) +
labs(x = "Depth", y = "Temperature", title = "Lake Temperatures by Depth") 
print(Lakes.Temp.by.depth)

```

