---
title: "Assignment 8: Time Series Analysis"
author: "Laurie Muzzy"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on time series analysis.

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio. 
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
You will need to have the correct software installed to do this (see Software Installation Guide)
Press the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Assignments folder.
8. After Knitting, please submit the completed exercise (PDF file) to the dropbox in Sakai. Please add your last name into the file name (e.g., "Salk_A08_TimeSeries.pdf") prior to submission.

The completed exercise is due on Tuesday, 19 March, 2019 before class begins.

## Brainstorm a project topic
1. Spend 15 minutes brainstorming ideas for a project topic, and look for a dataset if you are choosing your own rather than using a class dataset. Remember your topic choices are due by the end of March, and you should post your choice ASAP to the forum on Sakai.

Question: Did you do this?

> ANSWER: Yes! I'm going to look at the Pb datasets from EPA Outdoor Air Quality from Detroit, MI, from 1987-2017, to determine what sites have decreased in lead exposure over time.

## Set up your session 
2. Set up your session. Upload the EPA air quality raw dataset for PM2.5 in 2018, and the processed NTL-LTER dataset for nutrients in Peter and Paul lakes. Build a ggplot theme and set it as your default theme. Make sure date variables are set to a date format.

```{r}
getwd()
library(nlme)
library(lubridate)
library(multcompView)
library(lsmeans)
library(trend)
library(tidyverse)
library(tidyr)

EPAair_PM25_NC2018_raw <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv")
#View(EPAair_PM25_NC2018_raw)

EPAair_PM25_NC2018_raw$Date <- as.Date(EPAair_PM25_NC2018_raw$Date, 
                                               format = "%m/%d/%y")
class(EPAair_PM25_NC2018_raw$Date) #Date
EPAair_PM25_NC2018_raw$AQS_PARAMETER_DESC <- "PM2.5"

PeterPaul.chem <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv")
#View(PeterPaul.chem)

PeterPaul.chem$sampledate <- as.Date(PeterPaul.chem$sampledate, 
                                               format = "%Y-%m-%d")
class(PeterPaul.chem$sampledate)

LFM8theme <- theme_classic(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom")
theme_set(LFM8theme)
```


## Run a hierarchical (mixed-effects) model

Research question: Do PM2.5 concentrations have a significant trend in 2018?

3. Run a repeated measures ANOVA, with PM2.5 concentrations as the response, Date as a fixed effect, and Site.Name as a random effect. This will allow us to extrapolate PM2.5 concentrations across North Carolina.

3a. Illustrate PM2.5 concentrations by date. Do not split aesthetics by site.

```{r}
#3 
#PM2.5 = response
#Date = fixed effect    `correlation = structure(form = ~ time | subjvar)'
#Site.Name = random effect
EPAair_PM25_NC2018_raw = EPAair_PM25_NC2018_raw[order(EPAair_PM25_NC2018_raw[,'Date'],-EPAair_PM25_NC2018_raw[,'Site.ID']),] 
EPAair_PM25_NC2018_raw = EPAair_PM25_NC2018_raw[!duplicated(EPAair_PM25_NC2018_raw$Date),]

PM2.5mixed <- lme(data = EPAair_PM25_NC2018_raw, 
                  Daily.Mean.PM2.5.Concentration ~ Date,  # response ~ explan
                  random = ~1|Site.Name, #random
                  correlation = corAR1(value = 0.513, form = ~ Date|Site.Name),
                  method = "REML")

summary(PM2.5mixed) #AIC 1756.622; pval is kinda high, so it says DATE not a sig predictor 

PM2.5fixed <- gls(data = EPAair_PM25_NC2018_raw,
                  Daily.Mean.PM2.5.Concentration ~ Date,
                  method = "REML")
summary(PM2.5fixed) 

anova(PM2.5mixed, PM2.5fixed) 
  
#3a
PM2.5Site <- EPAair_PM25_NC2018_raw %>%
  select(Date, Daily.Mean.PM2.5.Concentration, Site.Name) %>%
  na.exclude()
#View(PM2.5Site)

PM2.5inNC <- ggplot(PM2.5Site, aes(x = Date, y = Daily.Mean.PM2.5.Concentration)) +
  geom_point(size = 0.5, alpha = 0.5, color = "brown") +
  labs(x = "Date", y = "PM2.5 Concentration, ug/m3")
print(PM2.5inNC)
```

3b. Insert the following line of code into your R chunk. This will eliminate duplicate measurements on single dates for each site.
PM2.5 = PM2.5[order(PM2.5[,'Date'],-PM2.5[,'Site.ID']),]
PM2.5 = PM2.5[!duplicated(PM2.5$Date),]

3c. Determine the temporal autocorrelation in your model. 

3d. Run a mixed effects model. 

```{r}
#3c temporal autocorrelation
PM2.5corr <- lme(data = EPAair_PM25_NC2018_raw, 
                  Daily.Mean.PM2.5.Concentration ~ Date,
                  random = ~1|Site.Name)
PM2.5corr
ACF(PM2.5corr) # ACF = 0.513

#3d mixed effects model
PM2.5mixed <- lme(data = EPAair_PM25_NC2018_raw, 
                  Daily.Mean.PM2.5.Concentration ~ Date,  # response ~ explan
                  random = ~1|Site.Name, #random
                  correlation = corAR1(value = 0.513, form = ~ Date|Site.Name),
                  method = "REML")
PM2.5mixed
```

Is there a significant increasing or decreasing trend in PM2.5 concentrations in 2018? 

> ANSWER: There isn't a significant trend in PM2.5 concentrations over the course of the year, evidenced from the ACF value of 0.51 (about 50% of the concentrations are correlated to the values of the day before or after, which makes sense).

3e. Run a fixed effects model with Date as the only explanatory variable. Then test whether the mixed effects model is a better fit than the fixed effect model. 

```{r}
PM2.5fixed <- gls(data = EPAair_PM25_NC2018_raw,
                  Daily.Mean.PM2.5.Concentration ~ Date)
summary(PM2.5fixed) 

anova(PM2.5mixed, PM2.5fixed)
#           Model df      AIC      BIC    logLik   Test  L.Ratio p-value
#PM2.5mixed     1  5 1756.622 1775.781 -873.3110                        
#PM2.5fixed     2  3 1865.202 1876.698 -929.6011 1 vs 2 112.5802  <.0001
```

Which model is better?

> ANSWER:  The AIC is lower in the mixed effects model, so MIXED is better.


## Run a Mann-Kendall test

Research question: Is there a trend in total N surface concentrations in Peter and Paul lakes? 

4. Duplicate the Mann-Kendall test we ran for total P in class, this time with total N for both lakes. Make sure to run a test for changepoints in the datasets (and run a second one if a second change point is likely). 

```{r}
PeterPaul.N.surface <- PeterPaul.chem %>%
  select(-lakeid, -depth_id, -comments) %>%
  filter(depth == 0) %>%
  filter(!is.na(tn_ug)) 

ggplot(PeterPaul.N.surface, aes(x = sampledate,y = tn_ug, color = lakename)) +
  geom_point() +
  scale_color_manual(values = c("magenta", "cyan")) +
  labs(x = "Date", y = "Total N, micrograms")

Peter.N.surface <- filter(PeterPaul.N.surface, lakename == "Peter Lake")
Paul.N.surface <- filter(PeterPaul.N.surface, lakename == "Paul Lake")

#Peter Lake
mk.test(Peter.N.surface$tn_ug) #pval v low, z = 7.29, a significant positive trend
pettitt.test(Peter.N.surface$tn_ug) #low pval, significant change point at 36, from 1993-05-26

mk.test(Peter.N.surface$tp_ug[1:35]) #pval 0.589 , z = 0.53 so no trend 
mk.test(Peter.N.surface$tp_ug[36:98]) #pval 0.00531, z = -2.78 means a bit of a negative trend, but insignificant

pettitt.test(Peter.N.surface$tn_ug[36:98]) #36+21=57
mk.test(Peter.N.surface$tp_ug[57:98]) #pval = 0.129, z = -1.51, insignificant negative trend from 1994-06-29 to 1999-08-16
 
#Paul Lake
mk.test(Paul.N.surface$tn_ug) #pval 0.72, z = -0.35, insignificant negative trend
pettitt.test(Paul.N.surface$tn_ug) #change point at 16, from 1991-08-26

mk.test(Paul.N.surface$tn_ug[1:15]) #pval = 0.0075, z = -2.67, insignificant negative trend
mk.test(Paul.N.surface$tn_ug[16:99]) #pval = 0.0274, z = 2.20, insignificant positive trend

pettitt.test(Paul.N.surface$tn_ug[16:99]) #16+36=52, 5-17-1992
mk.test(Paul.N.surface$tn_ug[52:99]) #pval = 0.197, z = -1.28, insignificant negative trend

```


What are the results of this test?

> ANSWER: for Peter Lake: z = 7.2927, p-value = 3.039e-13. Since the p-val is so low, we can reject the null, meaning that we see a trend. Since the z-score is not near zero, we can say that there is a positive trend over time, i.e., Total N is getting higher in Peter Lake.
>However, Paul Lake (pval 0.72, z = -0.35) is not like this: the p-val is high, the z-score is close to zero, so we can't be confident that there's any sort of trend in Paul Lake.

5. Generate a graph that illustrates the TN concentrations over time, coloring by lake and adding vertical line(s) representing changepoint(s).

```{r}
PeterPaul.N <- ggplot(PeterPaul.N.surface, aes(x = sampledate, y = tn_ug, color = lakename)) +
  geom_point() +
  geom_vline(xintercept = as.Date("1991-08-26"), color = "orange", lty = 2) +
  geom_vline(xintercept = as.Date("1993-05-26"), color = "navy", lty = 1) +
  scale_color_manual(values = c("orange", "navy")) +
  labs(x = "Date", y = "Total N, micrograms")
print(PeterPaul.N)
```

 